---
title: "Nuforc Sightings Analysis"
author: "Savannah"
date: "2025-09-26"
output:
  html_document:
    theme: journal
    df_print: paged
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# File Setup:

These are all the libraries needed for this analysis.

```{r}
library(readr)
library(tidyverse)
library(ggplot2)
library(stopwords)
library(RColorBrewer)
library(wordcloud)
library(tidytext)
```

# Task 1: Building the Shape Table

## Filtering and cleaning the data:

First, I will read in the data:
```{r, show_col_types = FALSE}
raw_df<- read_csv("nuforc_sightings.csv")
state_map <- read_csv("states.csv")
```
To filter the data, I used the states.csv for abbreviations of states. Additionally, some states mismatched to other countries so I corrected that in the filtering as well.

```{r}
USA_df <- raw_df %>% mutate(state = tolower(state), 
                            country = tolower(country),
                            shape = tolower(shape)) %>% 
  filter(state %in% state_map$state | country == 'usa') %>% 
  mutate(state = recode(state,"ohio" = "oh", 
                        "new york" = "ny", "montana" = "mt",
                        "west virginia" = "wv", "wisconsin" = "wi")) %>% 
  filter(state %in% state_map$state | state == "-" |
           state == "0" | state == "dc")
```

With this data, I standardized the shape column, removing all empty strings and turning them into "unknown".

```{r}
stand_shape <- USA_df %>% 
  mutate(shape = ifelse(is.na(shape) | shape == "", "unknown",shape))
```

## Creating the pivot table:

```{r}
shape_by_state <- stand_shape %>% group_by(state, shape) %>%
  summarise(count = n(), .groups = "drop") 

shape_state_pivot <- shape_by_state %>% pivot_wider(
  names_from = shape, values_from = count, values_fill = 0)
```

There are 22 shapes in total excluding unknown and other. California/ CA saw the most circular shapes.

# Task 2: PCA on the Shape Table

## Row-normalizing the table and PCA:
```{r}
#Row normalization
shape_state_norm <- shape_state_pivot %>% rowwise() %>%
  mutate(total = sum(c_across(-state))) %>%   
  mutate(across(-c(state, total), ~ .x / total)) %>%
  select(-total) %>%
  ungroup()

# Performing PCA:
pca_shape <- prcomp(shape_state_norm[, -1], center = TRUE, scale. = TRUE)
```
## Proportion of variance explained and scree plot:

```{r}
pve_shape <- (pca_shape$sdev)^2 / sum(pca_shape$sdev^2)

plot(pve_shape, type = "b", xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", main = "Scree Plot")
```

It does seem that the variety of UFO sightings can be boiled down to a few key patterns. Based on the scree plot, the proportion of variance explained by the first few principle components are enough to cover the majority of variance in the data.

## Scatterplot:
```{r}
pca_shape_df <- as.data.frame(pca_shape$x)

ggplot(pca_shape_df, aes(x = PC1, y = PC2)) +
  geom_point(color = "steelblue", size = 3) +
  labs(title = "Shape PC1 vs PC2",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal()
```

Based on the scatterplot, the majority of states seem to be in one cluster. There are two states that exhibit outlier tendencies.

## Examining the PCA rotation:

```{r}
rotation_tbl <- as.data.frame(pca_shape$rotation) %>%
  round(3) %>% .[,c(1,2)] %>% 
  rownames_to_column(var = "Variable")
print(rotation_tbl)
```

It seems as though the shapes "light", "sphere", and "disk" affect PC1 the most. For PC2, the shapes "cylinder", "cube", and "unknown" contributed the most.

# Task 3: Clean and Tokenize Summaries

## To clean up the summaries:
```{r}
Encoding(USA_df$summary) <- "UTF-8"
clean_summary <- USA_df %>% 
  mutate(summary = tolower(summary)) %>%
  mutate(summary = iconv(summary, from = "UTF-8", to = "ASCII", sub = "")) %>%
  mutate(summary = gsub("\\s+", " ", trimws(summary)))
```

## Making an array of words and removing punctuation:
```{r}
word_array <- clean_summary %>% mutate(words = strsplit(summary, split = " "))
all_words <- unlist(word_array$words) %>% gsub("[[:punct:]]", "", .)
```

## Frequency of each word:
```{r}
freq_word <- table(all_words) %>% as.data.frame(.)
colnames(freq_word) <- c("word", "frequency")
freq_word <- freq_word[order(-freq_word$frequency), ]
```
## Table and word cloud:

```{r}
print(head(freq_word,50))
```

I chose to visualize this via a table. It is a very large table, so I only printed the first 50 rows.

```{r}
top_100_unclean <- head(freq_word, 100)
wordcloud(words = top_100_unclean$word,
          freq = top_100_unclean$frequency,
          min.freq = 1,
          random.order = FALSE,
          colors = brewer.pal(8, "Dark2"))
```

The most frequent words are generally generic English words or generic words used to describe UFOs.

## Removing stopwords from all_words:
```{r}
all_words_clean <- all_words[!all_words %in% stopwords('en')]

clean_word <- table(all_words_clean) %>% as.data.frame(.)
colnames(clean_word) <- c("word", "frequency")
clean_word <- clean_word[order(-clean_word$frequency), ]
```

## Histogram of cleaned words:
```{r}
top_20 <- head(clean_word, 20)

ggplot(top_20, aes(x = reorder(word, frequency), y = frequency)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +  # Flips to horizontal bars
  labs(title = "Top 20 Most Frequent Words",
       x = "Words",
       y = "Frequency") +
  theme_minimal()
```

From the top frequent words after cleaning, we find the words "light"/"lights"/"bright", "sky", and "moving" as some of the top characterizations of UFO sightings

# Task 4: Building the Keywords Table and PCA

## Defining the dictionary of 100 words:
```{r}
top_100 <- clean_word %>%
  filter(nchar(as.character(word)) >= 3) %>%
  slice_max(order_by = frequency, n = 100)
```

## To make table of frequencies by state:
```{r}
top100_df <- data.frame(word = top_100$word)
tokenized_df <- clean_summary %>%
  select(state, summary) %>%
  unnest_tokens(word, summary) %>% 
  filter(word %in% top100_df$word)

word_state_pivot <- tokenized_df %>%
  count(state, word, sort = TRUE) %>% 
  pivot_wider(names_from = word, values_from = n, values_fill = 0)

print(word_state_pivot)
```




## Performing Task 2 for this table
### Row-normalizing the table:
```{r}
word_state_norm <- word_state_pivot %>% rowwise() %>%
  mutate(total = sum(c_across(-state))) %>%   
  mutate(across(-c(state, total), ~ .x / total)) %>%
  select(-total) %>%
  ungroup()
```


### Performing PCA:
```{r}
pca_word <- prcomp(word_state_norm[, -1], center = TRUE, scale. = TRUE)
```


### Proportion of variance explained & scree plot:
```{r}
pve_word <- (pca_word$sdev)^2 / sum(pca_word$sdev^2)
plot(pve_word, type = "b", xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", main = "Scree Plot")
```

Similar to the shape by state data, this word count by state data also seems to be boiled down to a few patterns/ principle components. The majority of the variance is described by the first few principle components. The proportion of variance defined by latter PCs dwindle.

### Scatterplot:
```{r}
pca_word_df <- as.data.frame(pca_word$x)

ggplot(pca_word_df, aes(x = PC1, y = PC2)) +
  geom_point(color = "steelblue", size = 3) +
  labs(title = "Word Count PC1 vs PC2",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal()
```

Similarly to the shape scatterplot, the majority of states seem to be in one cluster. There are again two states that seem to be outliers.

### PCA rotation:
```{r}
rotation_word_tbl <- as.data.frame(pca_word$rotation) %>%
  round(3) %>% .[,c(1,2)] %>% 
  rownames_to_column(var = "Variable")
print(rotation_word_tbl)
```

The words "bright", "light", "star", "across", "orb", and "blinking" do suggest correlation with the shapes that were observed to be of primary drivers in the previous PCA. In general, the shapes that were most similarly described in this PCA are light", "sphere", "unknown" and "disk". The shapes "rectangle" and "cube" are less accurately described by these top words.


