---
title: "HW5:Clustering"
author: "Savannah"
date: "2025-10-21"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Libraries Needed:

```{r, message=FALSE}
library(tidyverse)
library(cluster)
library(glue)
library(gridExtra)
```


#Task 1: Clusters and the Gap Statistic

## Data Loading:

From the data_generation.py file, I generated the necessary datasets according to the directions on the PDF. I will now proceed to load all datasets.

```{r, message=FALSE}
file_list <- list.files("derived_data", pattern = "*.csv", full.names = TRUE)

# Read all CSVs into a named list of data frames
df_list <- file_list %>%
  set_names(~ basename(.) %>% tools::file_path_sans_ext()) %>%
  map(read_csv)

n <- c(2:6)
side_length <- c(1,10,2:9)
side_length
```

## Clustergap Simulations:

For each dataset, I will run the cluster gap function. Then, I will visualize the results.
***NOT SURE IF NEEDED***
```{r}
estim_cluster <- c()
for(df_name in names(df_list)){
  df <- df_list [[df_name]]
  cg <- clusGap(df %>% as.matrix(), kmeans, nstart = 20, iter.max = 50, K.max = 10, B = 10)
  plot(cg, main = glue("Cluster Gap for Dataset {df_name}"), xlab = "Number of Clusters")
}
```

These plots show the Gap statistics by the number of clusters for each dataset generated. As the range for the possible number of clusters is iterated through, a gap statistic is calculated for each potential number of clusters. This gap statistic compares the within cluster dispersion in the data and compares it to that of randomized data. The number of clusters associated with the largest value of the gap statistic is the optimal number of clusters supported by the dataset. 

## Plotting Estimated Number of Clusters by Side Length
```{r}
D2 <- df_list[grepl("^df_2_", names(df_list))]
D3 <- df_list[grepl("^df_3_", names(df_list))]
D4 <- df_list[grepl("^df_4_", names(df_list))]
D5 <- df_list[grepl("^df_5_", names(df_list))]
D6 <- df_list[grepl("^df_6_", names(df_list))]

```

```{r}
estimate_clusters <- function(df_list, n_value) {
  results <- list()
  
  for (name in names(df_list)) {
    df <- df_list[[name]]
    
    # Drop label column if it exists
    if ("label" %in% names(df)) {
      df <- df %>% select(-label)
    }

    # Extract side_length from name (e.g., df_2_5 → 5)
    side_length <- as.integer(str_extract(name, "\\d+$"))
    
    # Run clusGap (you can increase B later)
    set.seed(123)
    
    gap <- clusGap(df, FUN = kmeans, K.max = 10, nstart = 20, iter.max = 50, B= 10)
    
    # Get optimal number of clusters using firstSEmax
    opt_nc <- maxSE(gap$Tab[,"gap"], gap$Tab[,"SE.sim"], method = "firstSEmax")
    
    # Store results
    results[[name]] <- tibble(
      dataset = name,
      n = n_value,
      side_length = side_length,
      optimal_num_clusters = opt_nc
    )
  }
  
  # Combine and return
  bind_rows(results)
}
```

```{r, results = 'hide'}
gap_D2 <- estimate_clusters(D2, n_value = 2)
gap_D3 <- estimate_clusters(D3, n_value = 3)
gap_D4 <- estimate_clusters(D4, n_value = 4)
gap_D5 <- estimate_clusters(D5, n_value = 5)
gap_D6 <- estimate_clusters(D6, n_value = 6)
```

Now, I will first plot each gap statistic, and then I want to combine all of these ggplots into one figure for simplicity. 
```{r}

# Create individual plots
p2 <- ggplot(gap_D2, aes(x = side_length, y = optimal_num_clusters)) +
  geom_point() + geom_line() + ggtitle("n = 2") +
  geom_hline(yintercept = 2, linetype = "dashed", color = "red")

p3 <- ggplot(gap_D3, aes(x = side_length, y = optimal_num_clusters)) +
  geom_point() + geom_line() + ggtitle("n = 3") +
  geom_hline(yintercept = 3, linetype = "dashed", color = "red")

p4 <- ggplot(gap_D4, aes(x = side_length, y = optimal_num_clusters)) +
  geom_point() + geom_line() + ggtitle("n = 4") +
  geom_hline(yintercept = 4, linetype = "dashed", color = "red")

p5 <- ggplot(gap_D5, aes(x = side_length, y = optimal_num_clusters)) +
  geom_point() + geom_line() + ggtitle("n = 5")+
  geom_hline(yintercept = 5, linetype = "dashed", color = "red")

p6 <- ggplot(gap_D6, aes(x = side_length, y = optimal_num_clusters)) +
  geom_point() + geom_line() + ggtitle("n = 6")+ 
  geom_hline(yintercept = 6, linetype = "dashed", color = "red")

grid.arrange(p2, p3, p4, p5, p6, ncol = 3)

```

From these plots, we find that for n = 2, the optimal number of clusters is predicted is reduced after side_length = 1. For n = 3, n = 4 and n = 6, the optimal number of clusters is reduced for side_length < 3. At n = 5, side_length < 4 is when the optimal number of clusters is reduced.

To save it separately as a figure:
```{r, results='hide'}
if (!dir.exists("figures")) {
  dir.create("figures")
}

png("figures/cluster_gap_grid.png", width = 1000, height = 800)

grid.arrange(p2, p3, p4, p5, p6, ncol = 2)

dev.off()
```



