---
title: "Notepad"
output: html_document
---

```{r, message=FALSE}
#libraries:

library(tidyverse)

# dataloading:

ade <- read.csv('source_data/ad_events.csv')
ads <- read.csv('source_data/ads.csv')
camp <- read.csv('source_data/campaigns.csv')
user <- read.csv('source_data/users.csv')
```

```{r}
# definitely want to merge ade with ads and camp
# need to correct target_interests in ads and interests in user
# could merge ade composite with user via id
# goal for derived data: have three datasets, one composite of all datasets using ade as backbone, one for user, one for just ads, ade and camp together
```

```{r}
# cleaning up ads and users interests variable with onehot encoding
user2 <- user %>%
  mutate(user_interests = strsplit(interests, ",\\s*")) %>%  # split by comma + optional space
  unnest(user_interests) %>%
  mutate(value = 1) %>%
  pivot_wider(
    names_from = user_interests,
    values_from = value,
    values_fill = 0
  )

user2
write.csv(user2, 'derived_data/user_cleaned.csv')
```

```{r}
ads2 <- ads %>%
  mutate(target_interests = strsplit(target_interests, ",\\s*")) %>%  # split by comma + optional space
  unnest(target_interests) %>%
  mutate(value = 1) %>%
  pivot_wider(
    names_from = target_interests,
    values_from = value,
    values_fill = 0
  )

write.csv(ads2, 'derived_data/ads_cleaned.csv')
```

```{r}
comp1 <- ade %>% left_join(ads2, by = 'ad_id' ) %>% left_join(camp, by = 'campaign_id')
write.csv(comp1, 'derived_data/composite_data1.csv')
```


Questions that I want answered after data derivation:
0) How clean is this data? NAs, encodings? outliers?
1) What does the distribution of age of users look like?
2) Do campaigns/ ads actually draw in people of their target interest? On a similar note, do ads/ campaigns draw in the gender they want to draw in?
3) How does budget factor into the campaign? does higher budget = more people of that interest interacting with ads? Computation of a ratio of budget per day???
4) How does summer, winter/ seasons differ in terms of ad interactions? are there differences between the seasons? *Would probably want to take a look at those seasons as a separate variable???
5) Relationship between platform, type of ad, and interactions and type of user
6) Relationship between type of user and time of day
7) Relationship between platform and location of user?